### 2016-12-02 discussionX.Rmd
ENH:  Final commit.

\textcolor{blue}{
In evaluating the segmentation results, there are limited methodologies described in the
literature for automatic segmentation of WMHs despite the numerous clinical studies}
[@Debette:2010aa;@Kloppenborg:2014aa] \textcolor{blue}{exploring the connections between WMHs and TBI.
However, we can extrapolate from similar application areas where such methodologies have a
much longer history of development such as white matter lesion segmentation in
multiple sclerosis (MS).  In a recent literature review [@Garcia-Lorenzo:2013aa], 47 papers
involving automated segmentation of white
matter lesions in MS.  The most widely-used data for algorithmic evaluation consists
of two cohorts from two different sites used in
the MS Lesion segmentation challenge associated with the international MICCAI 2008
conference.  For comparison, this challenge training data set
has a mean lesion load of 204 ($\pm$ 752) $mm^3$ per lesion (compared with
our mean lesion load of 33.8 ($\pm$ 55.4) voxels $\times$ 1.2 $mm^3$ per voxel =
40.56 ($\pm$ 66.48) $mm^3$ per lesion) and the resolution is
almost twice what is used in this study (i.e., 0.5 $\times$ 0.5 $\times$ 0.5).
Note that our relative volume difference numbers are comparable to the
relative volume difference between raters of 68 percent (cf Figure 5 (d))}
[@vanginneken:2007aa].

>>>>

\textcolor{black}{
In evaluating the segmentation results, there are limited methodologies described in the
literature for automatic segmentation of WMHs despite the numerous clinical studies}
[@Debette:2010aa;@Kloppenborg:2014aa] \textcolor{black}{exploring the connections between WMHs and TBI.
However, we can extrapolate from similar application areas where such methodologies have a
much longer history of development such as white matter lesion segmentation in
multiple sclerosis (MS).  In a recent literature review [@Garcia-Lorenzo:2013aa], 47 papers
involving automated segmentation of white
matter lesions in MS.  The most widely-used data for algorithmic evaluation consists
of two cohorts from two different sites used in
the MS Lesion segmentation challenge associated with the international MICCAI 2008
conference.  For comparison, this challenge training data set
has a mean lesion load of 204 ($\pm$ 752) $mm^3$ per lesion (compared with
our mean lesion load of 33.8 ($\pm$ 55.4) voxels $\times$ 1.2 $mm^3$ per voxel =
40.56 ($\pm$ 66.48) $mm^3$ per lesion) and the resolution is
almost twice what is used in this study (i.e., 0.5 $\times$ 0.5 $\times$ 0.5).
Note that our relative volume difference numbers are comparable to the
relative volume difference between raters of 68 percent (cf Figure 5 (d))}
[@vanginneken:2007aa].

### 2016-12-02 figureCaptions.Rmd
BUG:  Switch subject ids.
ENH:  Final commit.

__Figure 3:__  Representation of Stage 1 feature images for subject 01C1019.  The
FLAIR, T1-, and T2-weighted images are rigidly pre-aligned
 to the space of the T1 image.  The three modality images are then preprocessed
(N4 bias correction and adaptive denoising) followed by
application of standard ANTs brain extraction and $n$-tissue segmentation protocols using
the MMRR symmetric template and corresponding priors applied to the T1 image.
The feature images are then generated for voxelwise input to the RF model which results in
the voting maps illustrated on the right.  This gives a probabilistic classification of
tissue type.  Not shown are the probability and voting images for the brain stem and
cerebellum.

@@

__Figure 5:__ Evaluation measures for both Stages of the leave-one-out protocol of the described protocol in the Methods
section:  (a) sensitivity, (b) positive predictive value, (c) $F_1$ score, and (d) relative
volume difference.   These quantitative assessments are given for three quantile ranges
spanning the range of the manually-derived lesion volumes.    Overall improvement in all
three whole lesion-based measuers is seen as the
second Stage RF model is applied for all three quantile ranges.
The relative volume difference corresponding to the Stage 2 results tend to predict a
decreased predicted volume over the Stage 1 results.

>>>>

__Figure 3:__  Representation of Stage 1 feature images for subject 9.  The
FLAIR, T1-, and T2-weighted images are rigidly pre-aligned
 to the space of the T1 image.  The three modality images are then preprocessed
(N4 bias correction and adaptive denoising) followed by
application of standard ANTs brain extraction and $n$-tissue segmentation protocols using
the MMRR symmetric template and corresponding priors applied to the T1 image.
The feature images are then generated for voxelwise input to the RF model which results in
the voting maps illustrated on the right.  This gives a probabilistic classification of
tissue type.  Not shown are the probability and voting images for the brain stem and
cerebellum.

@@

__Figure 5:__ Evaluation measures for both Stages of the leave-one-out protocol of the described protocol in the Methods
section:  (a) sensitivity, (b) positive predictive value, (c) $F_1$ score, and (d) relative
volume difference.   These quantitative assessments are given for three quantile ranges
spanning the range of the manually-derived lesion volumes.    Overall improvement in all
three whole lesion-based measures is seen as the
second Stage RF model is applied for all three quantile ranges.
The relative volume difference corresponding to the Stage 2 results tend to predict a
decreased predicted volume over the Stage 1 results.

### 2016-12-02 figureCaptions.tex
BUG:  Switch subject ids.
ENH:  Final commit.

\textbf{Figure 3:} Representation of Stage 1 feature images for subject
01C1019. The FLAIR, T1-, and T2-weighted images are rigidly pre-aligned
to the space of the T1 image. The three modality images are then
preprocessed (N4 bias correction and adaptive denoising) followed by
application of standard ANTs brain extraction and \(n\)-tissue
segmentation protocols using the MMRR symmetric template and
corresponding priors applied to the T1 image. The feature images are
then generated for voxelwise input to the RF model which results in the
voting maps illustrated on the right. This gives a probabilistic
classification of tissue type. Not shown are the probability and voting
images for the brain stem and cerebellum.

@@

\textbf{Figure 5:} Evaluation measures for both Stages of the
leave-one-out protocol of the described protocol in the Methods section:
(a) sensitivity, (b) positive predictive value, (c) \(F_1\) score, and
(d) relative volume difference. These quantitative assessments are given
for three quantile ranges spanning the range of the manually-derived
lesion volumes. Overall improvement in all three whole lesion-based
measuers is seen as the second Stage RF model is applied for all three
quantile ranges. The relative volume difference corresponding to the
Stage 2 results tend to predict a decreased predicted volume over the
Stage 1 results.

>>>>

\textbf{Figure 3:} Representation of Stage 1 feature images for subject
9. The FLAIR, T1-, and T2-weighted images are rigidly pre-aligned to the
space of the T1 image. The three modality images are then preprocessed
(N4 bias correction and adaptive denoising) followed by application of
standard ANTs brain extraction and \(n\)-tissue segmentation protocols
using the MMRR symmetric template and corresponding priors applied to
the T1 image. The feature images are then generated for voxelwise input
to the RF model which results in the voting maps illustrated on the
right. This gives a probabilistic classification of tissue type. Not
shown are the probability and voting images for the brain stem and
cerebellum.

@@

\textbf{Figure 5:} Evaluation measures for both Stages of the
leave-one-out protocol of the described protocol in the Methods section:
(a) sensitivity, (b) positive predictive value, (c) \(F_1\) score, and
(d) relative volume difference. These quantitative assessments are given
for three quantile ranges spanning the range of the manually-derived
lesion volumes. Overall improvement in all three whole lesion-based
measures is seen as the second Stage RF model is applied for all three
quantile ranges. The relative volume difference corresponding to the
Stage 2 results tend to predict a decreased predicted volume over the
Stage 1 results.

### 2016-12-02 introX.Rmd
ENH:  Final commit.

\textcolor{blue}{White matter hyperintensities (WMHs) are foci of abnormally increased signal intensity seen within white matter regions within the cerebrum and brainstem on fluid attenuation inversion recovery (FLAIR) magnetic resonance imaging (MRI) sequences.
WMHs are a frequent finding following traumatic brain injury (TBI) and have been correlated with functional outcome and injury severity in both pediatric} [@Bigler:2013aa;@Smitherman:2016aa] \textcolor{blue}{and adult} [@Marquez-de-la-Plata:2007aa;@Moen:2014aa;@Ding:2008aa;@Pierallini:2000aa] \textcolor{blue}{populations.}

\textcolor{blue}{Further research involving WMHs has shown that regional distribution and volume of WMHs have been shown to possess prognostic value in the TBI patient} [@Pierallini:2000aa;@Weiss:2008aa;@Smitherman:2016aa;@Levin:1988aa]\textcolor{blue}{. Specifically, lesion volume in corpus callosum correlates with functional scores in the acute phase following injury, while lesion volume in frontal lobes correlates with scores at 1 year following injury} [@Pierallini:2000aa]\textcolor{blue}{.
Further, volume of FLAIR lesions within the corpus
callosum, brainstem, and thalamus in patients with severe TBI correlates
with Glasgow Outcome-Extended (GOS-E) scores---a numeric groupwise assessment used to
classify ``outcome'' in TBI patients where ``outcome'' refers to the
spectrum of possible prognoses from death to disability to recovery} [@Moen:2014aa]\textcolor{blue}{. Additionally,
in patients who are comatose following severe TBI
the regional distribution of FLAIR lesions within the pons, midbrain, hypothalamus, basal forebrain, parietal, temporal, occipital lobes, and insula along with the observation of grasping or chewing behavior are associated with poor outcome} [@Weiss:2008aa].

\textcolor{blue}{Despite the above findings which demonstrate that WMHs have potential prognostic value, they are not routinely employed as a diagnostic measure in clinical practice.  Performing a comprehensive manual counting of number and distribution of lesions in the clinical setting is simply not practical. The development of such lesion quantification approaches may allow for the practical inclusion of this type of information within routine radiological practice.  In this work, we present an automated framework for quantification of WMHs in
multi-modal MRI using the random forest machine learning technique.}

@@

The random forests framework [@breiman2001] is a popular machine learning technique that has demonstrated significant utility for supervised segmentation tasks (e.g.,
normal human brain segmentation [@yi2009]) and other computer vision applications
[@viola2005].
\textcolor{blue}{Random forest-based paradigms have been successfully employed in the
delineation of other neuropathologies} [@geremia2011;@Pustina:2016aa;@geremia2012;@bauer2012;@zikic2012;@Tustison:2015aa]
\textcolor{blue}{for both single and multi-modal acquisition protocols.}

@@

Random forests are conceptually straightforward [@breiman2001].
\textcolor{blue}{The basic component of
the random forest paradigm is the ``decision tree'' often represented by a flowchart or
graph where internal nodes represent ``tests'', or decisions, and the edges represent the outcome of those tests.  The final, or end, nodes represent the various classifications produced by traversal through the decision tree.  For the proposed application, individual
voxels (and their corresponding feature values) are introduced at the root of a particular
decision tree and traverse the edges and internal nodes ultimately ending up at one of
the classification nodes according to the tests at each internal node.  A single random
forest model will consist of many such trees (often refered to as an ``ensemble'').}
\textcolor{blue}{Although decision trees had been extensively studied, the success of employing collections of such weak learners for boosting machine learning performance
(e.g., AdaBoost)} [@schapire1990;@freund1997]
\textcolor{blue}{influenced the similarly styled conglomeration of decision trees into ``forests'' with randomized node optimization} [@ho1995;@amit1997].
\textcolor{blue}{Finally, Breiman} [@breiman2001]
\textcolor{blue}{improved accuracy by random sampling of training data (i.e., ``bagging'') resulting
in the current random forest technique applied here.}
\textcolor{blue}{As voxels and their feature values are ``pushed'' through each decision tree in the forest,  votes for each label are accumulated and converted to probability values for all classification possiblilities at
each voxel location.}

>>>>

\textcolor{black}{White matter hyperintensities (WMHs) are foci of abnormally increased signal intensity seen within white matter regions within the cerebrum and brainstem on fluid attenuation inversion recovery (FLAIR) magnetic resonance imaging (MRI) sequences.
WMHs are a frequent finding following traumatic brain injury (TBI) and have been correlated with functional outcome and injury severity in both pediatric} [@Bigler:2013aa;@Smitherman:2016aa] \textcolor{black}{and adult} [@Marquez-de-la-Plata:2007aa;@Moen:2014aa;@Ding:2008aa;@Pierallini:2000aa] \textcolor{black}{populations.}

\textcolor{black}{Further research involving WMHs has shown that regional distribution and volume of WMHs have been shown to possess prognostic value in the TBI patient} [@Pierallini:2000aa;@Weiss:2008aa;@Smitherman:2016aa;@Levin:1988aa]\textcolor{black}{. Specifically, lesion volume in corpus callosum correlates with functional scores in the acute phase following injury, while lesion volume in frontal lobes correlates with scores at 1 year following injury} [@Pierallini:2000aa]\textcolor{black}{.
Further, volume of FLAIR lesions within the corpus
callosum, brainstem, and thalamus in patients with severe TBI correlates
with Glasgow Outcome-Extended (GOS-E) scores---a numeric groupwise assessment used to
classify ``outcome'' in TBI patients where ``outcome'' refers to the
spectrum of possible prognoses from death to disability to recovery} [@Moen:2014aa]\textcolor{black}{. Additionally,
in patients who are comatose following severe TBI
the regional distribution of FLAIR lesions within the pons, midbrain, hypothalamus, basal forebrain, parietal, temporal, occipital lobes, and insula along with the observation of grasping or chewing behavior are associated with poor outcome} [@Weiss:2008aa].

\textcolor{black}{Despite the above findings which demonstrate that WMHs have potential prognostic value, they are not routinely employed as a diagnostic measure in clinical practice.  Performing a comprehensive manual counting of number and distribution of lesions in the clinical setting is simply not practical. The development of such lesion quantification approaches may allow for the practical inclusion of this type of information within routine radiological practice.  In this work, we present an automated framework for quantification of WMHs in
multi-modal MRI using the random forest machine learning technique.}

@@

The random forests framework [@breiman2001] is a popular machine learning technique that has demonstrated significant utility for supervised segmentation tasks (e.g.,
normal human brain segmentation [@yi2009]) and other computer vision applications
[@viola2005].
\textcolor{black}{Random forest-based paradigms have been successfully employed in the
delineation of other neuropathologies} [@geremia2011;@Pustina:2016aa;@geremia2012;@bauer2012;@zikic2012;@Tustison:2015aa]
\textcolor{black}{for both single and multi-modal acquisition protocols.}

@@

Random forests are conceptually straightforward [@breiman2001].
\textcolor{black}{The basic component of
the random forest paradigm is the ``decision tree'' often represented by a flowchart or
graph where internal nodes represent ``tests'', or decisions, and the edges represent the outcome of those tests.  The final, or end, nodes represent the various classifications produced by traversal through the decision tree.  For the proposed application, individual
voxels (and their corresponding feature values) are introduced at the root of a particular
decision tree and traverse the edges and internal nodes ultimately ending up at one of
the classification nodes according to the tests at each internal node.  A single random
forest model will consist of many such trees (often refered to as an ``ensemble'').}
\textcolor{black}{Although decision trees had been extensively studied, the success of employing collections of such weak learners for boosting machine learning performance
(e.g., AdaBoost)} [@schapire1990;@freund1997]
\textcolor{black}{influenced the similarly styled conglomeration of decision trees into ``forests'' with randomized node optimization} [@ho1995;@amit1997].
\textcolor{black}{Finally, Breiman} [@breiman2001]
\textcolor{black}{improved accuracy by random sampling of training data (i.e., ``bagging'') resulting
in the current random forest technique applied here.}
\textcolor{black}{As voxels and their feature values are ``pushed'' through each decision tree in the forest,  votes for each label are accumulated and converted to probability values for all classification possiblilities at
each voxel location.}

### 2016-12-02 methods.Rmd
ENH:  Final commit.

Figure 1 provides a graphical overview of the proposed workflow.  The major components
include offline generation of symmetric multimodal templates, the creation of
feature images from the training data which are then employed for statistical
prediction using a concatenated random forest framework.  This framework involves
the use of two random forest models where the outcome (i.e., the tissue
probability estimates) of the first RF model application,
which we denote as "Stage1", is used as input (along with the original set of
feature images) to a refinement RF model segmentation which we denote as "Stage 2".
Once these offline steps
are performed, a new (i.e., unsegmented) subject can then be processed using the proposed pipeline.

@@

Several measures have been employed in the literature for evaluating
automated white matter lesion segmentation involving such quantities as lesion load,
voxel-based overlap measures (such as the Dice similarity coefficient), and lesion-based measures.[@Garcia-Lorenzo:2013aa].  For this work, due to the relatively small-size distribution of the lesion load in our data cohort (see Table 1), we used
four lesion-based measures:  sensitivity, positive predictive value (i.e., precision), $F_1$ score, and relative volume difference.  The first three quantities are based on the number of false positives ($TN$, false negatives ($FN$), and true positives ($TP$) in terms of identified lesions.  It should be noted that the number of true negatives is not readily
incorporated into measures of accuracy as the quantity of "true negatives" (i.e., correctly identified normal brain tissue) would severely skew accuracy assessments. The $F_1$ score is an assessment of accuracy which takes into account both the sensitivity,

>>>>

Figure 1 provides a graphical overview of the proposed workflow.  The major components
include offline generation of symmetric multimodal templates, the creation of
feature images from the training data which are then employed for statistical
prediction using a concatenated random forest framework.  This framework involves
the use of two random forest models where the outcome (i.e., the tissue
probability estimates) of the first RF model application,
which we denote as "Stage 1", is used as input (along with the original set of
feature images) to a refinement RF model segmentation which we denote as "Stage 2".
Once these offline steps
are performed, a new (i.e., unsegmented) subject can then be processed using the proposed pipeline.

@@

Several measures have been employed in the literature for evaluating
automated white matter lesion segmentation involving such quantities as lesion load,
voxel-based overlap measures (such as the Dice similarity coefficient), and lesion-based measures.[@Garcia-Lorenzo:2013aa].  For this work, due to the relatively small-size distribution of the lesion load in our data cohort (see Table 1), we used
four lesion-based measures:  sensitivity, positive predictive value (i.e., precision), $F_1$ score, and relative volume difference.  The first three quantities are based on the number of false positives \textcolor{blue}{($FP$), true negatives ($TN$)},
false negatives ($FN$), and true positives ($TP$) in terms of identified lesions.  It should be noted that the number of true negatives is not readily
incorporated into measures of accuracy as the quantity of "true negatives" (i.e., correctly identified normal brain tissue) would severely skew accuracy assessments. The $F_1$ score is an assessment of accuracy which takes into account both the sensitivity,

### 2016-12-02 methodsX.Rmd
ENH:  Final commit.

MR images utilized for this initial report were acquired from a single scanner involved in the Chronic Effects of Neurotrauma Consortium’s (CENC) observational study (see Walker et al., this issue). Briefly, participants were Operation Iraqi Freedom/Operation Enduring Freedom (OIF/OEF) era Service Members and Veterans between the ages of 18-60 years with prior combat exposure and deployment(s). The feature images \textcolor{blue}{were derived from MR acquisitions of} 26 subjects aged 39.6 $\pm$ 8.1 years (range 28--58 years). Within this cohort, 24 (92%) were considered positive for TBI based upon the potential concussive events (PCE) interview process described in detail in Walker et al., this issue. \textcolor{blue}{All lesions were isolated in the white matter of the cerebrum.  Table 1 provides a descriptive statistical summary of the variation in lesion
load across the selected cohort.}

@@

\textcolor{blue}{The first author (J. R. S.) performed the manual WMH tracings for all 26
subjects.  J. R. S. is a radiologist certified by the American Board of Radiology, with a certificate of advanced qualification in vascular and interventional radiology, over 18 years of research experience in TBI, and 6 years of clinical imaging experience.  All
multi-modal MR dicom image slices were converted to the nifti file format} [^1].
\textcolor{blue}{All nifti image volumes for each subject were rigidly aligned to the T1 image of that subject using the ANTs software} [@Avants:2014aa] \textcolor{blue}{.  The normalized MRI volumes were then provided to
J. R. S. who traced each lesion using the ITK-SNAP tool} [@Yushkevich:2006aa]
\textcolor{blue}{
which has multi-image overlay capabilities for visualizing all modalities in all three
canonical views.}

@@

\textcolor{blue}{Figure 1 provides a graphical overview of the proposed workflow.  The major components
include offline generation of symmetric multimodal templates, the creation of
feature images from the training data which are then employed for statistical
prediction using a concatenated random forest framework.  This framework involves
the use of two random forest models where the outcome (i.e., the tissue
probability estimates) of the first RF model application,
which we denote as ``Stage1'', is used as input (along with the original set of
feature images) to a refinement RF model segmentation which we denote as ``Stage 2''.
Once these offline steps
are performed, a new (i.e., unsegmented) subject can then be processed using the proposed pipeline.}

![\textcolor{blue}{Workflow illustration for the proposed pipeline.  Processing of the multi-modal
input MRI for a single subject, using the multi-modal symmetric template, results in
the generation of the feature images.  These feature images are used as input to the
Stage 1 RF model producing the initial RF probability map estimates.  The Stage 1
voting maps, the original feature images, and the Stage 2 RF model result in the
final voting maps which includes the WMH probability estimate.  Note that the RF models
are constructed once from a set of training data which are processed using the
same feature-construction pipeline as the single-subject input MRI.}](Figures/wmhPipeline.png)

@@

\textcolor{blue}{Following} [@Tustison:2014ab] \textcolor{blue}{and} [@Tustison:2015aa], \textcolor{blue}{optimally derived templates
serve for both brain tissue segmentation (for the derivation of tissue prior probability maps) and generation of asymmetry feature images.
For this work we use the multi-modal data available from the public MMRR data set}
[@landman2011]\textcolor{blue}{.  We used all 46 multi-modal acquisitions of that study to produce a multi-modal template according to the procedure described in} [@Avants:2010aa] \textcolor{blue}{which
results in a mean (in terms of both shape and intensity) multivariate template representing the entire cohort.  Mid-canonical slices of the FLAIR, T1, and T2 components are illustrated in Figure 2.}

@@

\textcolor{blue}{
Crucial to these supervised segmentation approaches are the creation and selection of
``features'' as input (i.e., feature images constructed from the training data)
in conjunction with expertly identified structures of interest
(i.e., WMHs) for model construction.}  For the targeted application in this work,
tissue classification is performed at the voxelwise level.  In other words, each voxel
within the region of interest is sent through the ensemble of decision trees and receives
a set of classification votes from each tree thus permitting a regression or classification
solution.  Since this procedure is performed at the voxelwise level, intensity information
alone is insufficient for good segmentation performance due to the lack of spatial context.
For example, as pointed out in [@Neema:2009aa], higher intensities can be found at the
periventricular caps in normal subjects which often confounds automated lesion detection
algorithms.  Other potential confounds include MR signal inhomogeneity and noise.  Therefore,
even though machine learning and pattern recognition techniques are extremely powerful and
have significant potential, just as crucial to outcome is the creative construction and
deployment of salient feature images which we detail below.

@@

The segmentation probability images described above are used as feature images to
provide a spatial context for the random forest model prediction step.  Additional spatial contextual
feature images include the distance maps [@maurer2003] based on the csf, gray matter, and
deep gray matter images.  These latter images are intended to help distinguish white matter
hyperintensities from false positives induced by the partial voluming at the gray/white
matter interface.  A third set of images are based on the voxel location within the
space of the template.
\textcolor{blue}{Similar feature images were used in}
[@Anbeek:2004aa]
\textcolor{blue}{although, unlike the proposed framework, this previous work lacks
normalization to the standard
coordinate system provided by the template to dramatically improve spatial specificity
across all subjects.  To generate these images, the T1 image of each subject is
registered to the T1 template component using a B-spline variant} [@Tustison:2013ac]
\textcolor{blue}{of the well-known ANTs Symmetric Normalization (SyN) algorithm} [@Avants:2011ab].
\textcolor{blue}{Using the derived transforms, the template coordinate images are warped back to the space of the individual subject.}

@@

In previous brain tumor segmentation work [@Tustison:2015aa], it was demonstrated that a
concatenated supervised approach, whereby the prediction output from the first random forest
model serves as partial input for a second random forest model, can significantly improve
segmentation performance.  We do the same thing for the work described here where we employ two stacked random forests (or two "stages").  The Stage 1
feature images of the training data (as described previously) are used to construct the
Stage 1 model.  The training data Stage 1 features are then used to produce the voxelwise
"voting maps" (i.e., the classification count of each decision tree for each tissue label) via the Stage 1 random forest model.  All the Stage 1 features plus the Stage 1 voting maps
are used as input to the Stage 2 model.  In addition, we use the Stage 1 voting maps as
tissue priors (i.e., probabilistic estimates of the tissue spatial locations) for a second application of the $6$-tissue segmentation algorithm with an
additional Markov Random Field spatial prior (MAP-MRF) [@Avants:2011aa].
\textcolor{blue}{In order to maximize the spatial information for the $n$-tissue segmentation process following the voxelwise RF classification of Stage 1, we use
all three aligned preprocessed images for multivariate segmentation during the
second stage.}
The resulting seven posterior probability images constitute a third additional feature image set for Stage 2.

@@

In addition, similar to our previous offering [^2], we plan on creating
a self-encapsulated example to showcase the proposed methodology
\textcolor{blue}{which will also be available on github.}[^3]  The fact that
the data will also be made available through the Federal Interagency Traumatic Brain Injury Research (FITBIR) repository along
with the manual labelings will facilitate reproducibility on the part of the reader
as well as any interest in extending the proposed framework to other data sets.

@@

\textcolor{blue}{Several measures have been employed in the literature for evaluating
automated white matter lesion segmentation involving such quantities as lesion load,
voxel-based overlap measures (such as the Dice similarity coefficient), and lesion-based measures.}[@Garcia-Lorenzo:2013aa]\textcolor{blue}{.  For this work, due to the relatively small-size distribution} \textcolor{blue}{of the lesion load in our data cohort (see Table 1), we used
four lesion-based measures:  sensitivity, positive predictive value (i.e., precision), $F_1$ score, and relative volume difference.  The first three quantities are based on the number of false positives ($TN$, false negatives ($FN$), and true positives ($TP$) in terms of identified lesions.  It should be noted that the number of true negatives is not readily
incorporated into measures of accuracy as the quantity of ``true negatives'' (i.e., correctly identified normal brain tissue) would severely skew accuracy assessments. The $F_1$ score is an assessment of accuracy which takes into account both the sensitivity,}

@@

\textcolor{blue}{and the positive predictive value (PPV),}

@@

\textcolor{blue}{such that $F_1$ is given by}

@@

\textcolor{blue}{The relative volume difference is calculated for each of the true positive lesions using
the manual and predicted lesion volumes:}

@@

\textcolor{blue}{To illustrate how the performance of our framework varies with lesion size, we calculated
the above measures based on evenly split quantiles of the manual estimates of lesion volumes into 3 groups.  These three size ranges (in terms of the number of voxels) are:}

* \textcolor{blue}{Quantile 1: $[1-12)$,}

* \textcolor{blue}{Quantile 2: $[12-28)$, and}

* \textcolor{blue}{Quantile 3: $[28-551]$}.

\textcolor{blue}{These quantiles are used to showcase the performance (cf Figure 5).}

>>>>

MR images utilized for this initial report were acquired from a single
scanner involved in the Chronic Effects of Neurotrauma Consortium’s
(CENC) observational study (see Walker et al., this issue). Briefly,
participants were Operation Iraqi Freedom/Operation Enduring Freedom
(OIF/OEF) era Service Members and Veterans between the ages of 18-60
years with prior combat exposure and deployment(s). The feature images
\textcolor{black}{were derived from MR acquisitions of} 26 subjects aged
39.6 $\pm$ 8.1 years (range 28--58 years). Within this cohort, 24 (92%)
were considered positive for TBI based upon the potential concussive
events (PCE) interview process described in detail in Walker et al.,
this issue. \textcolor{black}{All lesions were isolated in the white
matter of the cerebrum.  Table 1 provides a descriptive statistical
summary of the variation in lesion load across the selected cohort.}

@@

\textcolor{black}{The first author (J. R. S.) performed the manual WMH
tracings for all 26 subjects.  J. R. S. is a radiologist certified by
the American Board of Radiology, with a certificate of advanced
qualification in vascular and interventional radiology, over 18 years of
research experience in TBI, and 6 years of clinical imaging experience.
All multi-modal MR dicom image slices were converted to the nifti file
format} [^1]. \textcolor{black}{All nifti image volumes for each subject
were rigidly aligned to the T1 image of that subject using the ANTs
software} [@Avants:2014aa] \textcolor{black}{.  The normalized MRI
volumes were then provided to J. R. S. who traced each lesion using the
ITK-SNAP tool} [@Yushkevich:2006aa] \textcolor{black}{ which has
multi-image overlay capabilities for visualizing all modalities in all
three canonical views.}

@@

\textcolor{black}{Figure 1 provides a graphical overview of the proposed
workflow.  The major components include offline generation of symmetric
multimodal templates, the creation of feature images from the training
data which are then employed for statistical prediction using a
concatenated random forest framework.  This framework involves the use
of two random forest models where the outcome (i.e., the tissue
probability estimates) of the first RF model application, which we
denote as ``Stage1'', is used as input (along with the original set of
feature images) to a refinement RF model segmentation which we denote as
``Stage 2''. Once these offline steps are performed, a new (i.e.,
unsegmented) subject can then be processed using the proposed pipeline.}

![\textcolor{black}{Workflow illustration for the proposed pipeline.
Processing of the multi-modal input MRI for a single subject, using the
multi-modal symmetric template, results in the generation of the feature
images.  These feature images are used as input to the Stage 1 RF model
producing the initial RF probability map estimates.  The Stage 1 voting
maps, the original feature images, and the Stage 2 RF model result in
the final voting maps which includes the WMH probability estimate.  Note
that the RF models are constructed once from a set of training data
which are processed using the same feature-construction pipeline as the
single-subject input MRI.}](Figures/wmhPipeline.png)

@@

\textcolor{black}{Following} [@Tustison:2014ab] \textcolor{black}{and}
[@Tustison:2015aa], \textcolor{black}{optimally derived templates serve
for both brain tissue segmentation (for the derivation of tissue prior
probability maps) and generation of asymmetry feature images. For this
work we use the multi-modal data available from the public MMRR data
set} [@landman2011]\textcolor{black}{.  We used all 46 multi-modal
acquisitions of that study to produce a multi-modal template according
to the procedure described in} [@Avants:2010aa] \textcolor{black}{which
results in a mean (in terms of both shape and intensity) multivariate
template representing the entire cohort.  Mid-canonical slices of the
FLAIR, T1, and T2 components are illustrated in Figure 2.}

@@

\textcolor{black}{ Crucial to these supervised segmentation approaches
are the creation and selection of ``features'' as input (i.e., feature
images constructed from the training data) in conjunction with expertly
identified structures of interest (i.e., WMHs) for model construction.}
For the targeted application in this work, tissue classification is
performed at the voxelwise level.  In other words, each voxel within the
region of interest is sent through the ensemble of decision trees and
receives a set of classification votes from each tree thus permitting a
regression or classification solution.  Since this procedure is
performed at the voxelwise level, intensity information alone is
insufficient for good segmentation performance due to the lack of
spatial context. For example, as pointed out in [@Neema:2009aa], higher
intensities can be found at the periventricular caps in normal subjects
which often confounds automated lesion detection algorithms.  Other
potential confounds include MR signal inhomogeneity and noise.
Therefore, even though machine learning and pattern recognition
techniques are extremely powerful and have significant potential, just
as crucial to outcome is the creative construction and deployment of
salient feature images which we detail below.

@@

The segmentation probability images described above are used as feature
images to provide a spatial context for the random forest model
prediction step.  Additional spatial contextual feature images include
the distance maps [@maurer2003] based on the csf, gray matter, and deep
gray matter images.  These latter images are intended to help
distinguish white matter hyperintensities from false positives induced
by the partial voluming at the gray/white matter interface.  A third set
of images are based on the voxel location within the space of the
template. \textcolor{black}{Similar feature images were used in}
[@Anbeek:2004aa] \textcolor{black}{although, unlike the proposed
framework, this previous work lacks normalization to the standard
coordinate system provided by the template to dramatically improve
spatial specificity across all subjects.  To generate these images, the
T1 image of each subject is registered to the T1 template component
using a B-spline variant} [@Tustison:2013ac] \textcolor{black}{of the
well-known ANTs Symmetric Normalization (SyN) algorithm}
[@Avants:2011ab]. \textcolor{black}{Using the derived transforms, the
template coordinate images are warped back to the space of the
individual subject.}

@@

In previous brain tumor segmentation work [@Tustison:2015aa], it was
demonstrated that a concatenated supervised approach, whereby the
prediction output from the first random forest model serves as partial
input for a second random forest model, can significantly improve
segmentation performance.  We do the same thing for the work described
here where we employ two stacked random forests (or two "stages").  The
Stage 1 feature images of the training data (as described previously)
are used to construct the Stage 1 model.  The training data Stage 1
features are then used to produce the voxelwise "voting maps" (i.e., the
classification count of each decision tree for each tissue label) via
the Stage 1 random forest model.  All the Stage 1 features plus the
Stage 1 voting maps are used as input to the Stage 2 model.  In
addition, we use the Stage 1 voting maps as tissue priors (i.e.,
probabilistic estimates of the tissue spatial locations) for a second
application of the $6$-tissue segmentation algorithm with an additional
Markov Random Field spatial prior (MAP-MRF) [@Avants:2011aa].
\textcolor{black}{In order to maximize the spatial information for the
$n$-tissue segmentation process following the voxelwise RF
classification of Stage 1, we use all three aligned preprocessed images
for multivariate segmentation during the second stage.} The resulting
seven posterior probability images constitute a third additional feature
image set for Stage 2.

@@

In addition, similar to our previous offering [^2], we plan on creating
a self-encapsulated example to showcase the proposed methodology
\textcolor{black}{which will also be available on github.}[^3]  The fact
that the data will also be made available through the Federal
Interagency Traumatic Brain Injury Research (FITBIR) repository along
with the manual labelings will facilitate reproducibility on the part of
the reader as well as any interest in extending the proposed framework
to other data sets.

@@

\textcolor{black}{Several measures have been employed in the literature
for evaluating automated white matter lesion segmentation involving such
quantities as lesion load, voxel-based overlap measures (such as the
Dice similarity coefficient), and lesion-based
measures.}[@Garcia-Lorenzo:2013aa]\textcolor{black}{.  For this work,
due to the relatively small-size distribution} \textcolor{black}{of the
lesion load in our data cohort (see Table 1), we used four lesion-based
measures:  sensitivity, positive predictive value (i.e., precision),
$F_1$ score, and relative volume difference.  The first three quantities
are based on the number of false positives \textcolor{blue}{($FP$), true negatives ($TN$)},
false negatives ($FN$), and true positives ($TP$) in terms of identified lesions.  It
should be noted that the number of true negatives is not readily
incorporated into measures of accuracy as the quantity of ``true
negatives'' (i.e., correctly identified normal brain tissue) would
severely skew accuracy assessments. The $F_1$ score is an assessment of
accuracy which takes into account both the sensitivity,}

@@

\textcolor{black}{and the positive predictive value (PPV),}

@@

\textcolor{black}{such that $F_1$ is given by}

@@

\textcolor{black}{The relative volume difference is calculated for each
of the true positive lesions using the manual and predicted lesion
volumes:}

@@

\textcolor{black}{To illustrate how the performance of our framework
varies with lesion size, we calculated the above measures based on
evenly split quantiles of the manual estimates of lesion volumes into 3
groups.  These three size ranges (in terms of the number of voxels)
are:}

* \textcolor{black}{Quantile 1: $[1-12)$,}

* \textcolor{black}{Quantile 2: $[12-28)$, and}

* \textcolor{black}{Quantile 3: $[28-551]$}.

\textcolor{black}{These quantiles are used to showcase the performance
(cf Figure 5).}
